---
title: "15 Seasons of LeBron vs. Jordan"
date: 2018-06-14
categories: ["R"]
tags: ["plots", "sports"]
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(scales)

datlj <- read_csv("../../data/jordan_v_james.csv") %>% 
  mutate(player = case_when(Tm == "MIA" ~ "LeBron",
                            Tm == "CLE" ~ "LeBron",
                            TRUE ~ "Jordan")) %>% 
  filter(Lg == "NBA") %>% 
  mutate_at(vars(G:PTS), as.numeric) %>%
  mutate(per_or_total = case_when(PTS > 100 ~ "total",
                                  TRUE ~ "per_game"))


total <- datlj %>% 
  filter(per_or_total == "total") %>% 
  group_by(player) %>% 
  mutate(`seasons played` = row_number(),
         cumulative_points = cumsum(PTS),
         cumulative_assists = cumsum(AST),
         cumulative_blocks = cumsum(BLK),
         cumulative_steals = cumsum(STL),
         cumulative_rebounds = cumsum(TRB),
         cumulative_threes = cumsum(`3P`),
         cumulative_turnovers = cumsum(TOV),
         cumulative_minutes = cumsum(MP)) %>% 
  ungroup()

totalj <- datlj %>% 
  filter(player == "Jordan") %>% 
  filter(per_or_total == "total") %>% 
  group_by(player) %>% 
  mutate(`seasons played` = row_number(),
         cumulative_points = cumsum(PTS),
         cumulative_assists = cumsum(AST),
         cumulative_blocks = cumsum(BLK),
         cumulative_steals = cumsum(STL),
         cumulative_rebounds = cumsum(TRB),
         cumulative_threes = cumsum(`3P`),
         cumulative_turnovers = cumsum(TOV),
         cumulative_minutes = cumsum(MP)) %>% 
  ungroup()

totall <- datlj %>% 
  filter(player == "LeBron") %>% 
  filter(per_or_total == "total") %>% 
  group_by(player) %>% 
  mutate(`seasons played` = row_number(),
         cumulative_points = cumsum(PTS),
         cumulative_assists = cumsum(AST),
         cumulative_blocks = cumsum(BLK),
         cumulative_steals = cumsum(STL),
         cumulative_rebounds = cumsum(TRB),
         cumulative_threes = cumsum(`3P`),
         cumulative_turnovers = cumsum(TOV),
         cumulative_minutes = cumsum(MP)) %>% 
  ungroup()
```


The confetti has barely settled on the Golden State Warriors' victory parade, and there are already people saying that the comparison is over: Michael Jordan is better than LeBron James at basketball. LeBron cannot be considered the Greatest Of All Time. Jordan left with six rings, while LeBron has only three.

My own feeling is that when the stats are this close, there is no objective way to pick a GOAT. We have a [missing data problem](https://en.wikipedia.org/wiki/Rubin_causal_model#The_fundamental_problem_of_causal_inference): it's impossible to see Jordan play on the 2005 Cavs, or watch LeBron go up against a team of space aliens. One must simply pick a side based on favorable stats from below, and then dig in.

Either way, now is an ideal time to compare the two. They have both played 15 seasons in the NBA. These are not advanced statistics - just a few season career totals of basic measures. The first, and arguably most important, is points. Here, Jordan edges out LeBron's `r comma(totall$cumulative_points[15])` with `r comma(totalj$cumulative_points[15])`

```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_points, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Points")
```

It's worth noting that while they played a similar number of minutes, Michael Jordan took a break for part of his 9th season to [play baseball](https://www.tarheelblog.com/2018/6/15/17462422/michael-jordans-break-unc-tarheels-baseball-nba-basketball-legacy-questions-goat). All of the comparisons have this blip in the data.  

```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_minutes, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Minutes")
```


Despite the difference in minutes, there is a clear winner in some of the individual categories.

```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_assists, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Assists")
```


```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_steals, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Steals")
```


```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_blocks, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Blocks")
```


```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_rebounds, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Rebounds")
```


```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_threes, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Threes")
```


```{r, echo=FALSE}
total %>% 
  ggplot(aes(x = `seasons played`, y = cumulative_turnovers, color = player)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values=c("#CC6666", "#000333")) +
  scale_y_continuous(label = comma) +
  ylab(NULL) +
  ggtitle("Cumulative Turnovers")
```


# Conclusion
LeBron is better. At rebounds, assists, and threes. Jordan wins at scoring, steals, and turnovers. Lebron has `r comma(totall$cumulative_blocks[15])` blocks and Jordan has `r comma(totalj$cumulative_blocks[15])`, so that is a little too close for comfort. Like I said, pick a side and dig in. 